**OptiClust: Improved method for assigning amplicon-based sequence data to operational taxonomic units**

### 2016-10-24 submission/Westcott_OptiClust_mSystems_2016.Rmd
First full draft of Discussion section

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
library(dplyr)
library(tidyr)
library(knitr)

@@

## Discussion
Restate most important contributions
* Optimized clustering based on an objective criteria. Results are meaningfully better than existing methods
* Added benefits include smaller CPU and RAM footprint
* Result is efficient analysis of large datasets without sacrificing clustering quality as has been experienced in heuristic methods.

Choice of objective criteria
* Value of using a metric that is based on all four parameters
* Preference is for Matthew's Correlation Coefficient because... other options include F1 Score and accuracy

Preference for OptiClust over cluster splitting approach
* Risks are mis-classification and artificially splitting similar sequences between OTUs because they classify to different taxa
* Still potential to merge the methods for more efficient processing.

Data quality - analysis uses unique sequences and we should not expect an infinite number of unique sequences unless there is a large amount of random sequencing error.

Based on our model, we propose the following...

>>>>

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
library(dplyr, warn.conflicts=FALSE)
library(tidyr)
library(knitr)

@@

## Discussion
Myriad methods have been proposed for assigning 16S rRNA gene sequences to OTUs that each claim improved performance based on speed, memory usage, representation of taxonomic information, and number of OTUs. Each of these metrics is subjective and do not actually indicate the quality of the clustering. This led us to propose using the MCC as a metric for assessing the quality of clustering, post hoc. Here, we described a new clustering method that seeks to optimize clustering based on an objective criterion that measures clustering quality, a priori. In the OptiClust algorithm clustering is driven by  optimizing a metric that assesses whether any two sequences should be grouped into the same OTU. The result are clusters that are significantly more robust and is efficient in the time and memory required to cluster the sequences into OTUs. This makes it more tractable to analyze large datasets without sacrificing clustering quality as was previously necessary using heuristic methods.

The cluster optimization procedure is dependent on the metric that is chosen for optimization. We employed the MCC because it includes the four values from a confusion matrix. Other algorithms such as the furthest neighbor and nearest neighbor algorithms minimize the number of FP and FN, respectively; however, these suffer because the number of FN and FP are not controlled. Alternatively, one could optimize based on the sensitivity, specificity, or accuracy, which are each based on two values from the confusion matrix or they could optimize based on the F1 score, which is based on three values from the confusion matrix. Because these metrics do not balance all four parameters equally, it is likely that one parameter will dominate in the optimization procedure. For example, optimizing for sensitivity could lead to a large number of FPs. Since we would like to minimize both FPs and FNs and not just the total number of false assignments, we decided to optimize utilizing the MCC. The MCC measures the correlation between observed and predicted classifications and is robust to cases where there is an uneven distribution across the confusion matrix. It is possible that other metrics could be developed and employed for optimization of the clustering.

The OptiClust algorithm is relatively simple. For each sequence it effectively ask whether the MCC value will increase if the sequence is moved to a different OTU including creating a new OTU. If the value does not change, it remains in the current OTU. The algorithm repeats until the MCC value stabilizes. Assuming that the algorithm is seeded with each sequence in a separate OTU, it does not appear that the algorithm converges to a local optimum. Furthermore, execution of the algorithm with different random number generator seeds produces OTU assignments of consistently high quality. Future improvements to the implementation of the algorithm could provide optimization to further improve its speed and susceptibility to find a local optimum. Users are encourage to repeat the OTU assignment several times to confirm that they have found the best OTU assignments.

Our previous MCC-based analysis of numerous clustering algorithms indicated that the average neighbor algorithm consistently produced the best OTU assignments with the DGC-based method using USEARCH also producing robust OTU assignments. The challenge in using the average neighbor algorithm is that it requires a large amount of RAM and is computationally demanding. This led to the development of a splitting approach that divides the clustering across distinct taxonomic groups. The improved performance provided by the OptiClust algorithm likely makes such splitting unnecessary for most current datasets. We have demonstrated that although the OTU assignments made at the genus level are still better than that of other methods, the quality is not as good as that found without splitting. The loss of quality is likely due to misclassification because of limitations in the clustering algorithms and reference databases. The practical significance of such small differences in clustering quality remain to be determined; however, based on the current analysis, it does appear that the number of OTUs is artificially inflated. Regardless, the best clustering quality should be pursued given the available computer resources.

The time required to execute the OptiClust algorithm scaled proportionally to the number of unique sequences raised to the second or third power and the amount of RAM required scaled to the second power. The power for the time requirement is affected by the similarity of the sequences in the dataset with datasets containing more similar sequences having a higher power. Also, the number of unique sequences is the basis for both the amount of time and memory required to complete the algorithm. Both the similarity of sequences and number of unique sequences can be driven by the sequencing error since any errors will increase the number of unique sequences and these sequences will be closely related to the perfect sequence. This underscores the importance of reducing the noise in the sequence data. If sequencing errors are not remediated and are relatively randomly distributed, then it is likely that the algorithm will require an unnecessary amount of time and RAM to complete.

The rapid expansion in sequencing capacity has demonstrated that database-independent assignment of sequences into OTUs is a critical step in the sequence analysis pipeline. Of course, the quality of the assignments is dependent on the quality of the underlying data. The ability to objectively optimize the clustering quality will significantly enhance the quality of those OTU assignments while making the time and memory requirements more efficient. The development of the OptiClust algorithm represents a significant advance that likely has numerous other applications.
